\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Edits

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \sdmTMB\ was the fastest across all mesh resolutions; however, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB.
\pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \R\ algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.

@@

While our focus is primarily in applications to biological data, the SPDE approach and functionality of \sdmTMB\ has applications in many other fields.
Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.

TODO add paragraph here on uniting functionality not available in any one package?

@@

% TODO: figure out where to put this paragraph
% TODO: rework this paragraph to be advantages of random fields GAMs and CAR
The GLMMs underpinning \sdmTMB\ models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
% The random fields in \sdmTMB\ are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
% The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\code{share\_range} argument).
% TODO: figure out where to put this paragraph
Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids...
Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004} (e.g., spatial variance and range---the distance at which points are effectively independent).

@@

Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, time-varying spatiotemporal covariance parameters \citep{ward2022}, and simulation from the parameter joint precision matrix.

There are several planned future additions to the \sdmTMB\ model structure.
A subset of features to be added in the future include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
(4) The ability to specify observation-level likelihoods to integrate different data types (e.g. REF).
(5) Conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature.
(6) Improved visualization methods and integration

>>>>

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \sdmTMB\ was the fastest across all mesh resolutions; however, \INLA/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \sdmTMB.
\pkg{mgcv} was most affected by mesh resolution.
Our test was restricted to one core and default \R\ algebra libraries.
% all packages could run faster with optimized libraries and parallel processing.
% Results with optimized math libraries on one core (\pkg{openBLAS}: \citet{openblas} and \pkg{PARDISO}: \citet{pardiso7.2a}) resulted in a \textasciitilde10\% speed increase for \sdmTMB\ and \pkg{inlabru} and a \textasciitilde7--9-fold speed increase for \pkg{mgcv}.

@@

While our examples focused on applications to ecological data, the SPDE approach and the functionality of \sdmTMB\ has applications in many other fields.
Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.

\textbf{TODO add paragraph here on uniting functionality not available in any one package?}

@@

% TODO: figure out where to put this paragraph
% TODO: rework this paragraph to be advantages of random fields GAMs and CAR
The GLMMs underpinning \sdmTMB\ models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} \citep{liaw2002}, \pkg{MaxEnt} \citep{phillips2006}, and most smooths in \pkg{mgcv} \citep{wood2017a}).
% The random fields in \sdmTMB\ are structured as MVN constrained by a Matérn covariance function \citep{matern1986}.
% The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) \citep{haskard2007}.
% The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to \(\sim 0.13\) \citep{lindgren2015}---can be shared among random fields or estimated separately (\code{share\_range} argument).
% TODO: figure out where to put this paragraph
Approaches such as conditional autoregressive models (CAR) require continuous spatial data to be aggregated to polygons or grids.
Areal data (data aggregated to polygon or grid level) may be analyzed using other spatial models, including conditional (CAR) autoregressive models \citep[e.g.,][]{verhoef2018}; \sdmTMB\ can also fit models with areal data if each polygon has an associated centroid.
A benefit of the geostatistical approach over CAR or similar models is that the parameters describing spatial covariance can be more easily interpreted \citep{wall2004}.

@@

Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, and time-varying spatiotemporal covariance parameters \citep{ward2022}.

There are several planned future additions to the \sdmTMB\ model structure.
A subset of features to be added include (1) multivariate models to allow for simultaneous modelling of data from multiple groups, (2) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base will be written in \R\ rather than \proglang{C++}, (3) inclusion of a dispersion formula similar to \pkg{glmmTMB},
(4) the ability to specify observation-level likelihoods to integrate different data types (e.g. REF),
(5) conditional autoregressive (CAR) models as an alternative to random fields for large datasets that are already areal in nature, and
(6) improved integration with existing visualization methods such as \pkg{ggeffects} \citep{ggeffects}.
